---
title: "Obstetrics_oahu"
format: html
editor: visual
---

```{r}
##libraries
library(tidyverse)
library(sf) 
library(here)
library(tigris)
library(mapview)
library(tidycensus)
library(arcgis)
library(dplyr)
library(osmdata)
library(ggplot2)
library(tibble)
```

## CMS file 

```{r}
#reading in cms data
cms <- read.csv(here("data", "cmshawaii.csv"),
                colClasses = "character")

oahu_ob <- cms %>%
  filter(STATE_CD == "HI",
         FIPS_CNTY_CD == "003",
         PRVDR_CTGRY_CD == "01",
         PGM_TRMNTN_CD == "00",
         OB_SRVC_CD %in% c("1", "2", "3"))
```

exploring oahu obstetric hospitals

note: explain why the cms file is a valid source for hospital obstetrics

```{r}
#claude assisted vis. for xploration purposes, not included in analysis (claude for column name meanings and tables)
oahu_ob <- oahu_ob %>%
  mutate(
    ob_service = case_when(OB_SRVC_CD == "1" ~ "Staff",
                           OB_SRVC_CD == "2" ~ "Under arrangement",
                           OB_SRVC_CD == "3" ~ "Staff and arrangement"),
    nursery    = if_else(NEONTL_NRSRY_SRVC_CD %in% c("1", "2", "3"), "Yes", "No"),
    nicu       = if_else(NEONTL_ICU_SRVC_CD %in% c("1", "2", "3"), "Yes", "No"),
    beds       = as.numeric(BED_CNT),
    cert_date  = as.Date(CRTFCTN_DT, format = "%Y%m%d")
  )

# One-row-per-hospital overview
oahu_ob %>%
  select(FAC_NAME, CITY_NAME, ob_service, nursery, nicu, beds, cert_date) %>%
  arrange(desc(beds))
```

Missing tripler, probably because its a military only hospital. Soo I need to add manually.

```{r}
tripler <- tibble(
  FAC_NAME  = "TRIPLER ARMY MEDICAL CENTER",
  ST_ADR    = "1 JARRETT WHITE RD",
  CITY_NAME = "HONOLULU",
  ZIP_CD    = "96859",
  BED_CNT   = "450"   
)

birth_hospitals <- oahu_ob %>%
  select(FAC_NAME, ST_ADR, CITY_NAME, ZIP_CD, BED_CNT) %>%
  bind_rows(tripler) %>%
  mutate(beds   = as.numeric(BED_CNT),
         source = if_else(FAC_NAME == "TRIPLER ARMY MEDICAL CENTER",
                          "Tripler", "CMS"))

birth_hospitals
```

## openstreetmap 

### for hospital location layer 

(alternative to arcgis.)

Oms features

```{r}
# List the features of OSM data
available_features()
```

```{r}
###creating oahu boundary box. then, pulling from openstreetmap specifically hospitals from the anmenitys
oahu_bb <- c(-158.3, 21.2, -157.6, 21.75)

osm_hosp <- opq(bbox = oahu_bb) %>%
  add_osm_feature(key = "amenity", value = "hospital") %>%
  osmdata_sf()

osm_hosp
```

The output was 12 polygons (the hospitals) which openstreetmap makes through the 203 points. But I dont need all thoe points. So I need to get rid of these outlined points and only have the reference points for hospitals left.

```{r}
osm_hosp <- osm_hosp %>%
  unique_osmdata()

osm_hosp
```

Now theres 3 points and 12 polygons. Basically on OSM, people can choose how they "draw" things. They can do the polygons, which are outlines of buildings, or they can do "points". Points are more common for restaurants inside malls, or smaller locations.
